---
title: "Getting started with hubEvals"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(width = 200)
```

`hubEvals` provides tools for scoring and evaluating
[hubverse](https://hubverse.io)-formatted model outputs, typically from a
forecasting hub. It bridges the hubverse
[`model_out_tbl`](https://docs.hubverse.io/en/latest/user-guide/model-output.html)
format and the [`scoringutils`](https://epiforecasts.io/scoringutils/)
package, so that an analyst can move from a tibble of model outputs plus
the matching
[oracle output](https://docs.hubverse.io/en/latest/user-guide/target-data.html#oracle-output)
target data to a tibble of scores in a single call.

This vignette walks through the main scoring workflows for each output type
supported by hubEvals, using example data from the
[`hubExamples`](https://hubverse-org.github.io/hubExamples/) package. The
example data come from the
[example-complex-forecast-hub](https://github.com/hubverse-org/example-complex-forecast-hub),
a small synthetic flu forecasting hub.

```{r, message = FALSE}
library(hubEvals)
library(dplyr)
```

## The input data

`hubEvals` consumes two tibbles:

* a
  **[`model_out_tbl`](https://docs.hubverse.io/en/latest/user-guide/model-output.html)**
  of model predictions (one row per prediction)
* an
  **[`oracle_output`](https://docs.hubverse.io/en/latest/user-guide/target-data.html#oracle-output)**
  of target data at the same task-id resolution (one row per observation;
  the `oracle_value` column holds the observed value)

`hubExamples::forecast_outputs` is a `model_out_tbl` that contains
predictions in the five hubverse output types currently supported by
hubEvals: `quantile`, `mean`, `median`, `pmf`, and `sample`. (The
hubverse spec also defines a `cdf` output type, evaluation of which is
not yet supported.)

```{r}
head(hubExamples::forecast_outputs)
```

`hubExamples::forecast_oracle_output` is the matching oracle output.

```{r}
head(hubExamples::forecast_oracle_output)
```

The main entry point is `score_model_out()`. It accepts a `model_out_tbl`
of a **single** output type at a time, transforms it into a `scoringutils`
forecast object, computes scores using metrics appropriate for that output
type, and (by default) summarizes the per-row scores by model.

> **Note:** While `model_out_tbl` must be filtered to a single `output_type`
> per scoring call, `oracle_output` can contain rows for multiple
> `output_type`s. The same full `oracle_output` can be passed to every
> scoring call.

## Quantile forecasts

Quantile forecasts are the most common output type in current hubverse
hubs. To score them, filter the `model_out_tbl` to `output_type == "quantile"`
and pass it to `score_model_out()`:

```{r}
quantile_out <- hubExamples::forecast_outputs |>
  dplyr::filter(output_type == "quantile")

quantile_scores <- score_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output
)
quantile_scores
```

```{r, include = FALSE}
quantile_forecast <- transform_quantile_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output
)
quantile_metric_list <- paste0(
  "`", names(scoringutils::get_metrics(quantile_forecast)), "`",
  collapse = ", "
)
```

By default, `score_model_out()` returns the metrics that `scoringutils`
considers appropriate for the given output type. For quantile forecasts,
these are `r quantile_metric_list`. See `?score_model_out` for short
descriptions of each metric and the defaults for the other output types.

To compute only a subset, name them via the `metrics` argument (more in
[Selecting a subset of metrics](#selecting-a-subset-of-metrics) below):

```{r}
score_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output,
  metrics = c("wis", "ae_median")
)
```

### Custom interval coverage levels

To request interval coverage at different levels, name them explicitly via
the `metrics` argument (covered in general in [Selecting a subset of
metrics](#selecting-a-subset-of-metrics) below). `interval_coverage_XX`
reports coverage of the **central XX% interval**, which is bounded by the
`(100 - XX) / 200` and `1 - (100 - XX) / 200` quantiles. Both bounds must
be present in the data.

A common gotcha: having the `0.95` quantile in the data is not enough to
compute `interval_coverage_95`. The 95% central interval is bounded by the
0.025 and 0.975 quantiles; the 0.95 quantile is the upper bound of the
**90%** central interval (paired with 0.05). The example hub provides the
0.05, 0.10, 0.25, 0.50, 0.75, 0.90, and 0.95 quantiles, so only the
following interval coverage levels are computable:

| `interval_coverage_XX` | Lower quantile | Upper quantile |
|---|---|---|
| `interval_coverage_50` | 0.25 | 0.75 |
| `interval_coverage_80` | 0.10 | 0.90 |
| `interval_coverage_90` | 0.05 | 0.95 |

```{r}
score_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output,
  metrics = c("wis", "interval_coverage_80")
)
```

## Point forecasts (mean and median)

The hubverse defines two point forecast output types: `mean` and
`median`. `score_model_out()` scores `mean` output types with squared
error (`se_point`) and `median` output types with absolute error
(`ae_point`).
The pairing follows from a property of these metrics
([Gneiting 2011](https://doi.org/10.1198/jasa.2011.r10138)):

- The expected squared error of a prediction against draws from a
  distribution F is minimised when the prediction equals the **mean**
  of F.
- The expected absolute error is minimised when the prediction equals
  the **median** of F.

Squared error therefore directly measures how close a `mean` forecast is
to the true mean of the data, and absolute error does the same for the
`median`. A mismatched pairing (e.g. scoring `mean` output types with
absolute error) would measure how close the model's reported mean is to
the true median rather than the true mean.

```{r}
mean_out <- hubExamples::forecast_outputs |>
  dplyr::filter(output_type == "mean")

score_model_out(
  model_out_tbl = mean_out,
  oracle_output = hubExamples::forecast_oracle_output
)
```

```{r}
median_out <- hubExamples::forecast_outputs |>
  dplyr::filter(output_type == "median")

score_model_out(
  model_out_tbl = median_out,
  oracle_output = hubExamples::forecast_oracle_output
)
```

## Probability mass function (pmf) forecasts

PMF forecasts assign a probability to each category of a categorical target.
The example hub has a `wk flu hosp rate category` target with four
categories: `low`, `moderate`, `high`, `very high`.

The categories carry a natural ordering (severity), so scoring rules that
exploit that ordering give a more informative picture than scoring rules
that treat the categories as unordered labels. `score_model_out()` switches
between nominal (no ordering) and ordinal (with ordering) scoring based on
whether the `output_type_id_order` argument is provided.

### Nominal scoring (no category order)

```{r}
pmf_out <- hubExamples::forecast_outputs |>
  dplyr::filter(output_type == "pmf")
```

```{r, include = FALSE}
pmf_nominal_forecast <- transform_pmf_model_out(
  model_out_tbl = pmf_out,
  oracle_output = hubExamples::forecast_oracle_output
)
pmf_nominal_metric_list <- paste0(
  "`", names(scoringutils::get_metrics(pmf_nominal_forecast)), "`",
  collapse = ", "
)
```

Without `output_type_id_order`, the forecast is treated as nominal. The
default metric set is `r pmf_nominal_metric_list`. Nominal scoring does
not use any ordering between categories: any wrong category is "equally"
wrong, weighted only by the probability placed on the true one.

```{r}
score_model_out(
  model_out_tbl = pmf_out,
  oracle_output = hubExamples::forecast_oracle_output
)
```

Every model scores `Inf`. This is the proper scoring rule behaving
correctly: each of these models assigns probability `0` to some category
that is then observed, giving `log_score = -log(0) = Inf` on that row,
and averaging propagates the `Inf` through to the model's overall score.
The result makes cross-model comparison impossible whenever any single
row hits a zero. A common workaround, used by FluSight
([Reich et al. 2019](https://doi.org/10.1371/journal.pcbi.1007486)), is
to cap `log_score` at a finite value before aggregation; FluSight
truncates at `-10` under the `log(p)` convention, which corresponds to
capping `log_score = -log(p)` at `10` in the scoringutils convention
used here. The cap is a domain judgement and the resulting score is no
longer a proper scoring rule. Apply the cap manually using
`summarize = FALSE` to expose per-row scores (covered in
[Summarising scores](#summarising-scores) below):

```{r}
score_model_out(
  model_out_tbl = pmf_out,
  oracle_output = hubExamples::forecast_oracle_output,
  summarize = FALSE
) |>
  dplyr::mutate(log_score = pmin(log_score, 10)) |>
  dplyr::group_by(model_id) |>
  dplyr::summarise(log_score = mean(log_score), .groups = "drop")
```

### Ordinal scoring (with category order)

```{r, include = FALSE}
pmf_ordinal_forecast <- transform_pmf_model_out(
  model_out_tbl = pmf_out,
  oracle_output = hubExamples::forecast_oracle_output,
  output_type_id_order = c("low", "moderate", "high", "very high")
)
pmf_ordinal_metric_list <- paste0(
  "`", names(scoringutils::get_metrics(pmf_ordinal_forecast)), "`",
  collapse = ", "
)
```

Providing `output_type_id_order` declares the order of the categories from
lowest to highest. `score_model_out()` then scores the forecasts as
ordinal, with the default metric set `r pmf_ordinal_metric_list`. Ordinal
scoring uses the declared category order, so a probability mass placed
one category away from the truth is penalised less than one placed
several categories away. A forecaster who is "nearly right" on the
severity scale will therefore score better under ordinal scoring than
under nominal scoring, which treats any wrong category as equally wrong.

The category order is declared in the hub's `tasks.json` config (under
`output_type.pmf.output_type_id.required` for the relevant model task).
For this hub, the order is `low`, `moderate`, `high`, `very high`:

```{r}
score_model_out(
  model_out_tbl = pmf_out,
  oracle_output = hubExamples::forecast_oracle_output,
  output_type_id_order = c("low", "moderate", "high", "very high")
)
```

`log_score` runs into the same `Inf` issue described above; `rps` is
finite even when the model assigns `0` probability to the observed
category, so it remains comparable across models without any correction.
The same manual cap workaround applies to `log_score` here.

If a hub has pmf targets with different bin sets (for example, one target
with three categories and another with five), score them in separate calls
so that each call has a single coherent category set.

## Sample forecasts

Sample-based forecasts represent the predictive distribution as a set of
draws. The hubverse supports two scoring modes:

* **Marginal**: each modeling task (each combination of task IDs) is scored
  independently. This is appropriate when the inferential target is the
  marginal predictive distribution at each task ID.
* **Compound**: samples are scored jointly across a subset of task ID
  dimensions, treating each sample draw as a joint prediction over those
  dimensions. This is appropriate when the inferential target is the joint
  distribution (for example, a trajectory over horizons).

### Marginal sample scoring

```{r, include = FALSE}
sample_out <- hubExamples::forecast_outputs |>
  dplyr::filter(output_type == "sample")
sample_marginal_forecast <- transform_sample_model_out(
  model_out_tbl = sample_out,
  oracle_output = hubExamples::forecast_oracle_output
)
sample_marginal_metric_list <- paste0(
  "`", names(scoringutils::get_metrics(sample_marginal_forecast)), "`",
  collapse = ", "
)
```

By default, `score_model_out()` scores sample forecasts marginally. The
default metric set is `r sample_marginal_metric_list`, headlined by the
[continuous ranked probability score (CRPS)](https://epiforecasts.io/scoringutils/reference/crps_sample.html).
Below we restrict the output to just CRPS for readability, using the
`metrics` argument covered in
[Selecting a subset of metrics](#selecting-a-subset-of-metrics) below:

```{r}
score_model_out(
  model_out_tbl = sample_out,
  oracle_output = hubExamples::forecast_oracle_output,
  metrics = "crps"
)
```

### Compound sample scoring

```{r, include = FALSE}
sample_compound_forecast <- transform_sample_model_out(
  model_out_tbl = sample_out,
  oracle_output = hubExamples::forecast_oracle_output,
  compound_taskid_set = c("reference_date", "location")
)
sample_compound_metric_list <- paste0(
  "`", names(scoringutils::get_metrics(sample_compound_forecast)), "`",
  collapse = ", "
)
```

To score samples jointly, supply `compound_taskid_set`: the task ID columns
that stay constant within a single sample draw. The remaining task ID
columns vary within a draw and are scored jointly using multivariate scoring
rules: the default metric set is `r sample_compound_metric_list`.

The `compound_taskid_set` is declared in the hub's `tasks.json` config
(under `output_type.sample.output_type_id_params.compound_taskid_set`).
For the example hub, each sample draw spans all horizons for a given
`reference_date` and `location`, so the `compound_taskid_set` is
`c("reference_date", "location")`:

```{r}
score_model_out(
  model_out_tbl = sample_out,
  oracle_output = hubExamples::forecast_oracle_output,
  compound_taskid_set = c("reference_date", "location")
)
```

`score_model_out()` validates that the submitted samples are compatible with
the requested `compound_taskid_set` and errors with a clear message if they
are not.

## Selecting a subset of metrics

The `metrics` argument applies across all output types: pass any subset of
the metric names valid for the relevant output type to compute only those.
For quantile forecasts, for example:

```{r}
score_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output,
  metrics = c("wis", "ae_median")
)
```

## Summarising scores

By default, `score_model_out()` returns one row per model, with each metric
averaged across all task-ID combinations. Two arguments control this
behaviour for any output type.

### Per-row scores

Setting `summarize = FALSE` returns one score per row of the input
`model_out_tbl`, with no aggregation across task-ID combinations:

```{r}
score_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output,
  summarize = FALSE
) |>
  head()
```

### Custom grouping

A custom `by` argument summarises by a different grouping, for example
each model's scores by horizon:

```{r}
score_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output,
  by = c("model_id", "horizon")
)
```

## Relative skill scores

Specifying `relative_metrics` adds a column expressing each model's score
for that metric relative to the rest of the model set. `relative_metrics`
should only be set to positive-valued, lower-is-better metrics. Proper
scoring rules like `wis`, `crps`, `log_score`, `rps`, `ae_point`,
`se_point`, `ae_median` are the natural choices; the WIS decomposition
components `overprediction`, `underprediction`, and `dispersion` also
work mathematically, though `dispersion` should be interpreted with care
(a smaller ratio means a sharper forecast, which is only beneficial if
the model is calibrated). Interval coverage metrics and `bias` are
rejected outright by `score_model_out()`: interval coverage targets a
nominal level rather than a "lower is better" direction, and `bias` is a
signed quantity for which the geometric mean of pairwise ratios is
undefined.

By default, the column is named `<metric>_relative_skill` and holds the
geometric mean of pairwise score ratios across all other models. Values
below 1 indicate a model that scored lower on average than its opponents.

```{r}
score_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output,
  metrics = c("wis", "interval_coverage_80", "interval_coverage_90"),
  relative_metrics = "wis"
)
```

Providing a `baseline` rescales the relative skill so that the chosen
baseline model has a skill of 1, and returns it in a column named
`<metric>_scaled_relative_skill`. Other models' values are then
interpretable directly with respect to the baseline (e.g. 0.7 means
"30% lower WIS than the baseline on average"):

```{r}
score_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output,
  metrics = c("wis", "interval_coverage_80", "interval_coverage_90"),
  relative_metrics = "wis",
  baseline = "Flusight-baseline"
)
```

Relative-skill computation requires that `"model_id"` is included in `by`,
which is the default.

## Scale transformations

Quantile, mean, median, and sample forecasts can be scored on a
transformed scale, which is useful when forecasts span several orders of
magnitude or when an exponential-growth process is better assessed on the
log scale. Pass a transformation function to the `transform` argument:

```{r}
score_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output,
  metrics = "wis",
  transform = scoringutils::log_shift,
  offset = 1
)
```

`scoringutils::log_shift()` is recommended for log transformations because
the `offset` argument handles zeros gracefully; using `log()` directly on
data that contain zeros produces `-Inf` and propagates `NaN` through the
scores.

To compare natural-scale and transformed-scale scores side by side, set
`transform_append = TRUE`. The result has a `scale` column distinguishing
rows scored on each scale:

```{r}
score_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output,
  metrics = "wis",
  transform = scoringutils::log_shift,
  transform_append = TRUE,
  offset = 1
)
```

### Custom transform labels

When `transform_append = TRUE`, the `scale` column labels the transformed
rows using a name inferred from the `transform` function (e.g.
`"log_shift"` for `scoringutils::log_shift`, `"sqrt"` for `sqrt`).
Override the inferred label by passing `transform_label` explicitly, for
example to shorten `"log_shift"` to `"log"`:

```{r}
score_model_out(
  model_out_tbl = quantile_out,
  oracle_output = hubExamples::forecast_oracle_output,
  metrics = "wis",
  transform = scoringutils::log_shift,
  transform_label = "log",
  transform_append = TRUE,
  offset = 1
)
```

For anonymous transformations, `transform_label` is **required**, because
there is no function name to infer from:

```{r}
score_model_out(
  model_out_tbl = mean_out,
  oracle_output = hubExamples::forecast_oracle_output,
  transform = function(x) sqrt(x),
  transform_label = "sqrt",
  transform_append = TRUE
)
```

Interval coverage metrics are invariant under monotonic transforms (the
0.25 quantile is still the 0.25 quantile after a log transform, so whether
the observation falls inside a central interval is unchanged), so
requesting them alongside a `transform` yields the same values on the
transformed scale as on the natural scale.

Scale transformations are not supported for `pmf` output types.

## Where to go next

* `score_model_out()` is a convenience that bundles the
  hubverse-to-scoringutils transformation with the scoring step. If you
  need the intermediate scoringutils forecast object (for example, to use
  `scoringutils` plotting helpers or to compose custom scoring metrics),
  the `transform_quantile_model_out()`, `transform_point_model_out()`,
  `transform_pmf_model_out()`, and `transform_sample_model_out()`
  exported helpers produce it directly.
* See `?score_model_out` for the full argument reference and a list of
  available metrics per output type.
* The
  [`scoringutils`](https://epiforecasts.io/scoringutils/) package provides
  rich tooling for visualizing and further analyzing the score tibbles
  produced by `score_model_out()`, including
  [`plot_wis()`](https://epiforecasts.io/scoringutils/reference/plot_wis.html)
  and
  [`plot_pairwise_comparison()`](https://epiforecasts.io/scoringutils/reference/plot_pairwise_comparison.html).
* The
  [`hubExamples`](https://hubverse-org.github.io/hubExamples/) package
  contains additional example data sets for experimenting with hubEvals.
